Copulas and their potential for ecology

### 2019-06-12 Paper.Rmd
edits to the Intro and master figure

The goal of this paper is to illustrate the potential for 
applications of copulas in ecology. 
Our results provide initial support for the following 
overall hypothesis: environmental, ecological and evolutionary processes 
generate and transmit copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. This is a broad hypothesis, and 
even the substantial empirical and theoretical
evidence we present here has only begun to test it. So we hope our work stimulates further
evaluation of the hypothesis by the research community.
We examine our overall hypothesis by assessing several more specific 
hypotheses, described below and outlined in Fig. 
\ref{MasterFigure}, all of which bear upon the overall hypothesis.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Guide for hypotheses addressed in the text and how each instantiates the
overall hypothesis. Labels are defined 
in the text to refer to parts of the figure. "Causal hypotheses" (i.e., hypotheses about 
causes of copula structure) are labeled with letters A-D. "Consequential 
hypotheses" (i.e., hypotheses about 
consequences of copula structure) are labeled with X-Z. The "observational 
hypothesis" (i.e., our hypothesis about 
copula structure of data) is labeled O.\label{MasterFigure}}
\end{center}
\end{figure}

One way we may expect a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portion of the distributions that dominate the association is visible in the copula
approach, but not via standard techniques. We therefore frame "causal hypothesis" A 
(labeled in Fig. \ref{MasterFigure}): 
copulas of dependencies between environmental and 
ecological variables will sensitively reveal 
nonlinear, asymmetric and threshold effects for population, community, 
and ecosystem functioning variables; copulas between 
environmental and ecological variables
will commonly show tail dependence; such tail dependence will lead to 
tail dependence across space in ecological variables.

A second reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of the spatial synchrony of population, community, 
biogeochemical and other ecological variables.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017]. We therefore frame 
causal hypothesis B (again labeled in Fig. \ref{MasterFigure}): Moran effects
transmit copula structure in spatial dependence from environmental to 
ecological variables, and hence copula structure and tail dependence will be 
common features of spatial synchrony in ecological variables.

A third reason we may expect a copula approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
would suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F suggests that Sp2
is dominant. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables, and this 
information can help support or refute alternative causal hypotheses. We
frame causal hypothesis C (Fig. \ref{MasterFigure}): copulas of abundances
through time or space of two competing or otherwise interacting species will
reflect any asymmetries in the interaction. We will later
frame causal hypothesis D, which concerns character evolution.

Another reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We therefore frame "consequential hypothesis" X (labeled on Fig. \ref{MasterFigure}):
details, revealed by copulas, of the 
dependence between local quantities can influence the spatial 
mean or sum, and this may influence higher organizational levels in 
ecology and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have more benign effects. We will
later frame consequential hypotheses Y, about extinction risk, and Z, about Taylor's law.

Our causal and consequential hypotheses imply, or rely on, an "observational
hypothesis" (O in Fig. \ref{MasterFigure}), which we here frame. Datasets in 
ecology will commonly have copula structure
distinct from that of a Gaussian/normal distribution (here called 
non-normal copula structure); they will commonly show tail dependence 
distinct from that of a normal distribution; in particular they will
show asymmetric tail dependence.

Because the observational hypothesis bears upon all the causal
and consequential hypotheses, we test it first, in sections 2-4 of the paper.
We analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
Sections 5 and 6 test the causal hypotheses, and sections 7 and 8 test the
consequential hypotheses. Section 9 is the Discussion.
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, these concepts have never been used in ecology. 
We argue that benefits to ecology of using a complete 
copula description of dependence will be substantial.

>>>>

The goal of this paper is to explore the potential for 
applications of copulas in ecology. 
Our results provide support for the idea that environmental, ecological and evolutionary processes 
generate and transmit nontrivial copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. The work we present here has begun to test this
hypothesis, and we hope our results stimulate further
evaluation of the hypothesis by the research community.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Summary and guide for analyses presented in the text. Middle boxes
correspond to ecological datasets for which Q1 was examined (in the 
second through
fourth sections of the text, see also Table \ref{tab_data_info}). Upper boxes 
correspond to potential causes of non-normal (NN) copula structure that
were examined (in the fifth and sixth sections of the text). Lower boxes 
correspond to potential consequences we examined of NN copula structure for
ecological understanding and for applications (in the seventh and eighth
sections of the text). Arrow labels (A-D for analyses pertaining 
to causes, Q2; and X-Z for analyses pertaining to consequences, Q3) 
link to locations in the text.\label{MasterFigure}}
\end{center}
\end{figure}

One way we may expect, *a priori*, a copula approach 
to better reflect ecological relationships relates to Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portion of the distributions that dominate the association is visible in the copula
approach, but not via standard techniques.

A second reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect copula structure in environmental variables 
to take the form of strong tail dependence if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of the spatial synchrony of population, community, 
biogeochemical and other ecological variables.
Synchrony is a phenomenon of great interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A third reason we may expect a copula approach to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
would suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F suggests that Sp2
is dominant. The copula approach again reveals which portions of the
distributions are principally responsible for the association between variables, and this 
information can help support or refute alternative causal hypotheses.

A fourth reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, captured by copulas, of the dependence between local
quantities can influence the spatial mean or sum, and how this may influence 
higher organizational levels in eoclogy and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have copula structure distinct from that of a Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of normal distribution. Such 
approaches do not utilize the power of copulas, so Q1 asks whether ecological
data contain dependency information distinct from what can be modelled with
standard approaches. (Q2) If the answer to
Q1 is "yes," what are some possible causes/mechanisms of non-normal
copula structure? And (Q3) what are the consequences of non-normal copula
structure and tail dependence for our understanding of ecology and for 
applications? To address Q1 (second through fourth sections of the paper), 
we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
The fifth and sixth sections of the paper 
address Q2 using simple models. The seventh and eigths sections address
Q3 using both data and models. The ninth section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, non-normal copulas have never been used in ecology. 
We argue that benefits to ecology of using a complete 
copula description of dependence will be substantial.
